sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-07-21 doc/paper-peerj.Rmd
Edit abstract

---
title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
preprint: false
author:
  - name: Sean C. Anderson
    affiliation: 1
    corresponding: true
    email: sean.anderson@dfo-mpo.gc.ca
  - name: Eric J. Ward
    affiliation: 2
  - name: Philina A. English
    affiliation: 1
  - name: Lewis A. K. Barnett
    affiliation: 3
affiliation:
  - code: 1
    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - code: 2
    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
  - code: 3
    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
abstract: >
  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
  Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
  The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models). Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
  sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. We anticipate its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
output:
  bookdown::pdf_book:
    template: template-peerj.tex
    keep_tex: true
    pandoc_args: "--highlight=haddock"
bibliography: refs.bib
link-citations: yes
csl: peerj.csl
header-includes:
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
---

>>>>

---
title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
preprint: false
author:
  - name: Sean C. Anderson
    affiliation: 1
    corresponding: true
    email: sean.anderson@dfo-mpo.gc.ca
  - name: Eric J. Ward
    affiliation: 2
  - name: Philina A. English
    affiliation: 1
  - name: Lewis A. K. Barnett
    affiliation: 3
affiliation:
  - code: 1
    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - code: 2
    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
  - code: 3
    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
abstract: >
  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to. Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
  Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
  The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models). Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
  sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. We anticipate its application to the full spectrum of problems involving geostatistical data---both within and beyond ecology---will prove equally fruitful.
output:
  bookdown::pdf_book:
    template: template-peerj.tex
    keep_tex: true
    pandoc_args: "--highlight=haddock"
bibliography: refs.bib
link-citations: yes
csl: peerj.csl
header-includes:
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
---
